---
title: "How Often Do People Get Pulled Over for Speeding in North Carolina?"
author: ''
date: '2021-02-17'
slug: exploratory-analysis-of-north-carolina-traffic-stop-data
categories:
- eda
- traffic stops
- Highcharts
- data visualization
tags: []
subtitle: Visualizing speeding stops in North Carolina
summary: ''
authors: []
lastmod: '2021-02-17T19:08:49-05:00'
featured: no
image:
  caption: <span>Photo by <a href="https://unsplash.com/@introspectivedsgn?utm_source=unsplash&amp;utm_medium=referral&amp;utm_content=creditCopyText">Erik
    Mclean</a> on <a href="https://unsplash.com/s/photos/police-car?utm_source=unsplash&amp;utm_medium=referral&amp;utm_content=creditCopyText">Unsplash</a></span>
  focal_point: ''
  preview_only: no
projects: []
links:
- icon: github
  icon_pack: fab
  name: Github
  url: https://github.com/shanejorr/shanes-site/blob/main/content/post/2021-02-17-exploratory-analysis-of-north-carolina-traffic-stop-data/index.en.Rmd
draft: yes
---

```{r setup, echo = F}
knitr::opts_chunk$set(echo = F,
                      message = F,
                      warning = F,
                      error = F)
```

```{r importLibraries}
library(glue)
library(gt)
library(vroom)
library(tidycensus)
library(lubridate)
library(highcharter)
library(scales)
library(leaflet)
library(tigris)
library(widgetframe)
library(tidyverse)
```

```{r}
# create ggplot theme for this post
post_theme <- theme_minimal() +
  theme(legend.position = 'bottom',
        plot.title = element_text(size = 11),
        plot.subtitle = element_text(size = 10),
        axis.title = element_text(size = 10))

theme_set(post_theme)

# standard credits to all to all high charts plots

# credits to be used on all plots
add_credits <- function(hc) {
  hc %>%
    hc_credits(
      enabled = TRUE,
      text = "Data source: The Stanford Open Policing Project | US Census Bureau Pop. Estimates",
      href = FALSE
  )
}
```

We all have preconceptions about cops and speeding tickets. That small town is a speed trap. There are more cops out on holidays like the 4th of July and Memorial Day, just waiting to pull people over. Don't get a red sports car because you are more likely to get a speeding ticket. But, do we know if any of these preconceptions are true?

[The Stanford Open Policing Project](https://openpolicing.stanford.edu/) built a repository of traffic stop data. The repository contains data from over 40 states. One of the states with the most complete data is my home state, North Carolina, where there are data from over 24 million police stops. 

We'll take a first look at the NC data in this post. Focusing on stops for speeding, we'll look at which counties and cities have the most speeding stops per 100 residents and which days see the most stops. We may not answer any of the pressing questions from the introduction, but this post will give us a starting point in answering deeper questions in later posts.

```{r importData, cache = T}
nc_stops <- vroom('/home/shane/Documents/nc_statewide_2020_04_01.csv.zip',
                  col_select = -contains('raw'))

# get 2009 - 2013 5 year and aggregate
pop_years <- seq(2009, 2013, 1)

n_years <- length(pop_years)

nc_stops <- nc_stops %>%
  filter(str_detect(reason_for_stop, "^Speed")) %>%
  mutate(year = year(date)) %>%
  # only keep between years 2009 and 2013 
  # 2013 is final full year of stop data
  filter(year %in% !!pop_years)
```

```{r}
nc_stops <- nc_stops %>%
  # remove string so that it matches with census data
  mutate(county_name = str_remove_all(county_name, ' County'),
         # rename misspelled county name
         county_name = str_replace_all(county_name, 'Tyrell', 'Tyrrell'))
```

```{r}
# census and shapefile information ----------------

# recode integers fo dates to years
recode_date <- c(
  `3` = 2010,
  `4` = 2011,
  `5` = 2012,
  `6` = 2013,
  `7` = 2014,
  `8` = 2015
)

# import population data by county and race
county_pop <- get_estimates(geography = "county", state = 'NC',
                            product = 'population', 
                            time_series = T,
                            year = 2018) %>%
  # remove string so that it matches with census data
  # only keep 2010 - 2014
  filter(DATE %in% seq(3, 6)) %>%
  # remove string so that it matches with traffic stop data
  mutate(NAME = str_remove_all(NAME, ' County, North Carolina')) %>%
  # make density and population different columns
  pivot_wider(id_cols = c('NAME', 'DATE', 'GEOID'),
              names_from = 'variable', values_from = 'value') %>%
  # aggregate 2010 - 2014 so it can be merged with aggregated stop data
  # sum population so we can divide by total stops between 2010-2014
  group_by(NAME, GEOID) %>%
  summarize(POP = sum(POP),
            DENSITY = mean(DENSITY))

```

## Which counties have the most speeding stops?

```{r}
# per capita number of stops ----------------

# calculate number of stops overall by year
per_capita_stops <- nc_stops %>%
  filter(!is.na(county_name)) %>%
  group_by(county_name) %>%
  count() %>%
  arrange(county_name) %>%
  left_join(county_pop, 
            by = c('county_name' = 'NAME')) %>%
  # calculate stops per 100 people
  mutate(stops_per = n/(POP/100)) %>%
  # round stops in new column for tooltip
  mutate(stops_tooltip = round(stops_per, 1)) %>%
  # create additional data points for tooltip
  mutate(year_pop = number((POP/!!n_years), accuracy = 1, big.mark = ','),
         year_stops = number((n/!!n_years), accuracy = 1, big.mark = ','))
```

We'll start by looking at speeding stops per 100 residents in each county. Mathematically, this is the number of stops for speeding within in a county divided by the county's population, times 100. In calculating the number of stops, all departments are included: state troopers, county sheriffs, and local police departments. As long as the stop occurred in the county, it's counted. Data from 2009 to 2013 - the last year of complete data - are aggregated. Values represent the average yearly stops per 100 residents among these four years.

### Map of stops per county

The map below shows each county's speeding stops per 100 residents. No associations pop out from the map. Rates are not higher along the cost than in the mountains, I-40 counties do not have higher rates, border counties do not have higher rates, etc. Nothing stands out.

```{r}
# title and tooltip for stops per 100 plots
stops_per_title <-   "North Carolina County Speeding Stops Per 100 Residents: 2009 - 2013"

stops_per_tooltip <- "<b>{point.county_name}</b><br>
                      Stops per 100 residents: <b>{point.stops_tooltip}</b><br>
                      Population: <b>{point.year_pop}</b><br>
                      Stops per year: <b>{point.year_stops}</b>"

# chloropath map of speeding stops per 100 residents by NC county
h_county_cloro <- hcmap(
  "countries/us/us-nc-all",
  data = per_capita_stops,
  value = "stops_per",
  joinBy = c("fips", "GEOID"),
  dataLabels = list(enabled = FALSE),
  borderColor = "#FAFAFA",
  borderWidth = 0.1,
  tooltip = list(
      headerFormat = "",
      pointFormat = stops_per_tooltip
    )
  ) %>%
  hc_title(text = stops_per_title) %>% 
  hc_subtitle(text = "Hover over county to see county name and rate") %>%
  add_credits()

frameWidget(h_county_cloro, height = 400, width = '95%')
```

### Bar chart of stops per county

The bar chart below lets us get a better feel for the distribution in values between counties. We see a large amount of variation. Looking at both extremes in the distribution, Camden's stops per 100 residents is over 10 times the state rate and Mitchell's is less than a tenth of the state rate. 

Part of the inter-county differences are explained by random variation. And for this factor, [we would expect](http://www.stat.columbia.edu/~gelman/research/published/allmaps.pdf) counties with smaller populations to have more random variation. Because of this, it's not a surprise that small-population counties typically make up the extreme values on both tails of the distribution. However, Camden's is reeeaaaalllly extreme. Unfortunately, answering the why question is beyond this post's scope.

```{r}
# calcualte state-level rate to add to bar chart
nc_avg_stops_per <- per_capita_stops %>%
  ungroup() %>%
  summarize(
    n = sum(n),
    POP = sum(POP),
    stops_per = n/(POP/100)
  ) %>%
  select(stops_per) %>%
  .[[1]]


# bar chart of county stops per 100 residents
h_couty_bar <- per_capita_stops %>%
  arrange(desc(stops_per)) %>%
  hchart(., "bar", hcaes(x = county_name, y = stops_per)) %>%
    hc_xAxis(labels = list(enabled = FALSE),
             title = list(text = 'North Carolina counties')) %>%
    hc_yAxis(title = list(text = "Stops per 100 residents"),
              # add north carolina average as vertical line
             plotLines = list(
               list(label = list(text = "NC stops per 100", verticalAlign = 'middle', textAlign = 'left'), 
                    color = "#7F7F7F", width = 2, zIndex = 5,
                    value = nc_avg_stops_per, dashStyle = 'Dash'))
             ) %>%
    hc_tooltip(
      headerFormat = "",
      pointFormat = stops_per_tooltip
    ) %>%
  hc_title(text = stops_per_title) %>% 
  hc_subtitle(text = "Hover over bars to see county name") %>%
  add_credits()

frameWidget(h_couty_bar, height = 400, width = '95%')
```

### Relationship between county population density and number of stops

To further look into the relationship between county populations and the number of stops, the next plot highlights the association between each county's population density (x-axis) and the number of stops per resident (y-axis). However, there is no association. If you were to overlay a single fit line on the points, that line would be horizontal.

Now, there is more variation in low-population density counties. But again, this is expected because the lower populations bring more randomness.

```{r}
h_county_scatter <- per_capita_stops %>%
  mutate(density_tooltip = round(DENSITY, 0)) %>%
  hchart(., "scatter", hcaes(x = DENSITY, y = stops_per)) %>%
    hc_xAxis(type = 'logarithmic') %>%
    hc_tooltip(
        headerFormat = "",
        pointFormat = stops_per_tooltip
      ) %>%
    hc_title(text = "Relationship between county population density and per capita speeding stops") %>% 
    hc_subtitle(text = "Each dot is a North Carolina county | Data reflects 2009 - 2013") %>%
    hc_xAxis(title = list(text = 'County population density')) %>%
    hc_yAxis(title = list(text = "Number of speeding stops per 100 residents")) %>%
    add_credits()

frameWidget(h_county_scatter, height = 300, width = '95%')
```

## Which cities have the most speeding stops?

We'll now move from counties to cities and towns. The data set only includes the stop's county, it does not record the city. But, the department is included. For example, Winston-Salem Police Department is one department. To look at stops per city, we'll calculate the number of stops each town or city police department made. Thus, our city aggregates will not include stops by state troopers and county sheriffs. Therefore, we're looking not at which cities have the highest number of stops per 100 residents, but which local police departments make the most stops per 100 residents.

Only cities with over 5,000 people are included. And of course, we only include cities where we have both population and traffic stop data.

```{r}
# create city-level stop and population dataset

# get city populations

place_pop <- map_df(pop_years, function(yr) {
    get_acs(year = yr,
            survey = "acs5", 
            geography = "place", state = "NC", 
            variables = "B01003_001") %>%
    mutate(year = !!yr)
}) %>%
  # only keep cities with five years of population data
  group_by(GEOID) %>%
  mutate(n = n()) %>%
  filter(n == !!n_years) %>%
  # sum all five years of populations
  # we can do this because traffic stops will be summed for the entire five years
  group_by(GEOID, NAME) %>%
  summarize(pop = sum(estimate)) %>%
  # only keep towns with at least 7,500 people
  ungroup() %>%
  filter((pop/!!n_years) > 5000) %>%
  # clean up city names so they match traffic stop names
  mutate(NAME = str_remove_all(NAME, ", North Carolina"),
         NAME = str_remove_all(NAME, " town| city| CDP| village"),
         NAME = str_replace_all(NAME, "Charlotte", "Charlotte-Mecklenburg"))

# get department data
dept_stops <- nc_stops %>%
  # only keep local police departments
  filter(!is.na(department_name),
         str_detect(department_name, 'Police Dep')) %>%
  filter(!str_detect(department_name, 'College|University'),
         !str_detect(department_name, '^NC|^UNC'),
         !str_detect(department_name, 'Hospital')) %>%
  # clean up department names to match population dataset
  mutate(department_name = str_remove_all(department_name, " Police Department"),
         department_name = str_trim(department_name)) %>%
  # aggregate number of stops by department
  group_by(department_name) %>%
  count() %>%  
  # merge in population data
  left_join(place_pop, by = c('department_name' = 'NAME')) %>%
  # only keep departments where we have popualtion data
  drop_na(pop) %>%
  mutate(year_pop = number((pop/!!n_years), accuracy = 1, big.mark = ','),
         year_stops = number((n/!!n_years), accuracy = 1, big.mark = ','),
         stops_per = n/(pop/100),
         stops_per_tooltip = round(stops_per, 1))
```

### Map of stops per city by local police departments

The map below shows the number of speeding stops per 100 city residents for each police department. The circle size corresponds to the number of speeding stops per 100 residents, with larger circles representing more stops per 100 residents. I'm not sure if there is anything of interest in terms of spatial relationships, but readers can draw their own conclusions.

```{r}
# get place data
nc_places <- places("North Carolina", progress_bar = FALSE) %>%
  # only keep places that are in the stop dataset
  filter(GEOID %in% !!dept_stops$GEOID)

# merge place data with stop dataset
dept_stops_geo <- dept_stops %>%
  left_join(nc_places, by = 'GEOID') %>%
    mutate(across(starts_with("INTPT"), ~as.numeric(.)))
```

```{r}
# create leaflet map of places

# popup label
city_labels <- labs <- lapply(seq(nrow(dept_stops_geo)), function(i) {
  paste0( '<b>', dept_stops_geo[i, "department_name"], '</b><br>', 
          'Stops per 100 residents: <b>',  dept_stops_geo[i, "stops_per_tooltip"], '</b><br>',
          'Population: <b>', dept_stops_geo[i, "year_pop"], '</b><br>',
          'Stops per year: : <b>', dept_stops_geo[i, "year_stops"], '</b>') 
})

l <- dept_stops_geo %>%
  rename(lng = INTPTLON, lat = INTPTLAT) %>%
  leaflet() %>% 
  setView(lng = -79.0193, lat = 35.7596, zoom = 7) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addCircleMarkers(
   radius = ~stops_per,
   label = lapply(city_labels, htmltools::HTML),
   stroke = FALSE, fillOpacity = 0.3
  )

frameWidget(l, height = '400')
```

### Bar chart of stops per city

The next graph shows the distribution of stops per 100 residents by local police departments. Like the county distributions, there is a large amount of variation. It's tempting to tell ourselves just-so stories to explain the variation and those stories might even be correct. For example, Aberdeen and Jacksonville have two of the three highest rates. That's gotta be those military kids, right? We'll this could be true. But, further analysis is needed to justify the story.

```{r}
# plot stops per 100 for each city

# title and tooltip for stops per 100 plots
stops_per_title <-   "North Carolina City Speeding Stops Per 100 Residents: 2009 - 2013"
stops_per_tooltip <- "<b>{point.department_name}</b><br>
                      Stops per 100 residents: <b>{point.stops_per_tooltip}</b><br>
                      Population: <b>{point.year_pop}</b><br>
                      Stops per year: <b>{point.year_stops}</b>"

h_city_bar <- dept_stops %>%
  arrange(desc(stops_per)) %>%
  hchart(., "bar", hcaes(x = department_name, y = stops_per)) %>%
    hc_xAxis(labels = list(enabled = FALSE),
             title = list(text = 'North Carolina cities')) %>%
    hc_yAxis(title = list(text = "Stops per 100 residents")) %>%
    hc_tooltip(
      headerFormat = "",
      pointFormat = stops_per_tooltip
    ) %>%
  hc_title(text = stops_per_title) %>% 
  hc_subtitle(text = "Hover over bars to see city name") %>%
  add_credits()

frameWidget(h_city_bar, height = 450, width = '95%')
```

## Days with the highest number of stops

Finally, let's look at which days have the highest number of speeding stops in the graph below. Memorial Day, 4th of July, and Labor day show spikes in the number of speeding stops. Actually, the Memorial Day spike spans the whole weekend, from Friday to Monday. And the Thanksgiving jump also incorporates Black Friday. Finally, there is an additional spike in late March, which could be spring break related.

As an important caveat, these spikes do not tell us whether we are more likely to get pulled over on the days with spikes. The spikes could occur for two reasons: (1) more people are out driving, and / or (2) police place a greater emphasis on pulling over speeders. We would need data on the number of drivers each day to disentangle the two reasons. And that might be another blog post.

```{r}
# create dataset of daily aggregate number of of stops for all NC

daily_stops <- nc_stops %>%
  filter(year == max(year)) %>%
  # get daily totals for each interstate group
  drop_na(date) %>%
  count(date) %>%
  # calculate percentage of total number of stops
  ungroup() %>%
  mutate(total_stops = sum(n),
         perc_total = n / total_stops) %>%
  arrange(date)

# create abbreviated month and day for tooltip
daily_stops$date_tooltip <- glue("{month.abb[month(daily_stops$date)]} {day(daily_stops$date)}, {year(daily_stops$date)}")
```

```{r}
# line plot of daily stops
h_county_daily <- hchart(daily_stops, 'line', hcaes(x = date, y = n)) %>% 
  hc_xAxis(type = "datetime") %>%
  hc_tooltip(
    headerFormat = "",
    pointFormat = "<b>{point.date_tooltip}</b>
                      <br>Number of stops: <b>{point.n}</b>"
  ) %>%
  hc_title(text = 'Number Of Speeding Stops Per Day in North Carolina - 2013') %>% 
  hc_yAxis(title = list(text = '# of stops')) %>%
  hc_xAxis(title = list(text = NULL)) %>%
  # add lines and bands for dates
  hc_xAxis(
    plotLines = list(
      list(label = list(text = "July 4th"), color = "#7F7F7F", width = 2, zIndex = 1,
           value = datetime_to_timestamp(as.Date('2013-07-04')), dashStyle = 'Dash'),
      list(label = list(text = "Christmas"), color = "#7F7F7F", width = 2, zIndex = 2,
           value = datetime_to_timestamp(as.Date('2013-12-25')), dashStyle = 'Dash'),
      list(label = list(text = "Labor Day"), color = "#7F7F7F", width = 2, zIndex = 3,
           value = datetime_to_timestamp(as.Date('2013-09-02')), dashStyle = 'Dash'),
      list(label = list(text = "Thanksgiving"), color = "#7F7F7F", width = 2, zIndex = 4,
           value = datetime_to_timestamp(as.Date('2013-11-28')), dashStyle = 'Dash'),
      list(label = list(text = "Memorial Day"), color = "#7F7F7F", width = 2, zIndex = 5,
           value = datetime_to_timestamp(as.Date('2013-05-27')), dashStyle = 'Dash')
    )
  ) %>%
  add_credits()

frameWidget(h_county_daily, height = 300, width = '95%')
```