---
title: Bayesian Mediation Analysis
author: Shane Orr
date: '2020-11-25'
slug: bayesian-mediation-analysis
categories:
  - bayesian analysis
  - mediation analysis
tags: []
subtitle: "Using Bayesian methods to conduct mediation analysis R"
summary: ''
authors: []
lastmod: '2020-11-25T18:38:02-05:00'
featured: no
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []
links:
- icon: github
  icon_pack: fab
  name: Github
  url: https://github.com/shanejorr/shanes-site/blob/main/content/post/2020-11-25-bayesian-mediation-analysis/index.en.Rmd
  
draft: true
---

```{r setup, echo = F}
knitr::opts_chunk$set(echo = T,
                      message = F,
                      warning = F,
                      error = F)
```

```{r loadPackages}
library(rstanarm)
library(mediation)
library(tidybayes)
library(glue)
library(knitr)
library(ggeffects)
library(tidyverse)

# stan options
# detect and use max number of cores
options(mc.cores = parallel::detectCores())

# create ggplot theme for this post
post_theme <- theme_minimal() +
  theme(legend.position = 'bottom',
        plot.title = element_text(size = 10),
        axis.title = element_text(size = 10))

theme_set(post_theme)
```


```{r customFunctions}
relabel_plot_color <- scale_fill_brewer(palette = "Set2", type = "Qual", labels = c(`0`= "Working", `1` = "Not working"))

posterior_diff <- function(mod, new_df, effect_type) {
  
  # create one column dataframe that is the difference in linear predictions for working and non-working students
  # parameters:
  #   mod: model
  #   new_df: dataframe to create predictions from
  #   effect_type: type of effect (indirect, direct, total)

  mod_post <- posterior_linpred(mod, newdata = new_df, transform = T,
                              seed = 1837, n = 500) %>%
    as.data.frame()
  
  col_split <- ncol(mod_post)/2
  
  mod_diff <- unlist(mod_post[1:col_split] - mod_post[(col_split+1):ncol(mod_post)])
  
  diff <- data.frame(diff = mod_diff,
                     type_effect = effect_type)
  
  return(diff)
}

posterior_diff_plot <- function(post_diff, x_limits, plot_title)  {
  
  # plot posterior distribution of difference in working and no-working predictions
  # parameters:
  #   post_diff: dataframe created from posterior_diff with predicted posterior 
  #     difference between workers and non-workers
  #   x_limits: plot x limits
  #   plot_title: title of plot
  
  ggplot(post_diff, aes(x =diff)) +
    stat_halfeye(alpha = .7) +
    scale_x_continuous(limits = x_limits,
                       labels = scales::percent_format(accuracy = 1)) +
    labs(title = plot_title,
         x = 'Difference in probability of passing class\nNot working - working',
         y = 'Density',
         fill = NULL)
}

two_group_plot <- function(df, plot_title) {
  
  # plot posterior distribution of both workers and non-workers
  # parameters:
  #   df: dataframe for plotting; will be linear predictions for workers and non-workers
  #   plot_title: title of plot
  
  ggplot(df, aes(x = .value, fill = as.factor(work))) +
    stat_halfeye(alpha = .8) +
    scale_x_continuous(labels = scales::percent_format(accuracy = 1)) +
    relabel_plot_color +
    labs(title = plot_title,
         x = 'Probability of passing class',
         y = 'Density',
         fill = NULL)
}
```

## What is mediation analysis?

### A simple example

### A simple overview

## How do I conduct one using Bayesian methods?

```{r createDataset}
# create simulated dataset --------

# number of rows in dataset
n <- 500

# 1 if student works, 0 otherwise
work_order <- c(1, 0)
working <- rep(work_order, times = n/2)

# save for future use
work_string <- c('Working', 'Not working')

# how much student studies
# if working, number with mean of 22 and sd of 5
# if not working, mean of 17, sd of 5
set.seed(2938)
study_hours <- rnorm(n, 
                     mean = 25 + (working*5),
                     sd = 10)
study_hours_std <- (study_hours - mean(study_hours)) / (sd(study_hours) / 2)

final_grade_formula <- (.05 + .015*study_hours + .04*working)
final_grade <- rnorm(n, mean = final_grade_formula, sd = .05)
final_grade <- ifelse(final_grade > 1, 1, final_grade)
final_grade <- ifelse(final_grade < 0, 0, final_grade)
final_grade <- rbinom(n = n, size = 1, prob = final_grade)

grades <- data.frame(
  work = working,
  study_std = study_hours_std,
  study = study_hours,
  grade = final_grade
)
```

### Total effect the old fashioned way

```{r deMod}
# total effect model
total_effect <- stan_glm(grade ~ work, 
                           family = binomial('logit'),
                           seed = 938,
                           data = grades)
```

```{r tePlotPosterior, fig.height = 3, fig.width = 4}
total_effect %>%
  add_fitted_draws(newdata = data.frame(work = work_order), seed = 93, n = 300) %>%
  two_group_plot('Prob. of Passing based on Work Status')
```

```{r tePlotDiff, fig.height = 2.5, fig.width = 4}
diff_limits <- c(-.1, .35)

total_effect_post <- total_effect %>%
  posterior_diff(new_df = data.frame(work = work_order), effect_type = 'Total Effect')

posterior_diff_plot(total_effect_post, x_limits = diff_limits, plot_title = 'Total Effect of Working on Passing')
```

### Direct effect the old fashioned way

```{r deMod}
full_mod <- stan_glm(grade ~ study + work, 
                     family = binomial('logit'),
                     seed = 938,
                     data = grades)

full_mod_newdata <- expand_grid(study = grades$study,
                                work = work_order) %>%
  arrange(desc(work), study) %>%
  as.data.frame()
```

```{r iePlotDiff, fig.height = 2.5, fig.width = 4}
direct_effect_post <- full_mod %>%
  posterior_diff(new_df = full_mod_newdata, effect_type = 'Direct Effect')

posterior_diff_plot(direct_effect_post, x_limits = c(-.15, .3), plot_title = 'Direct Effect of Working on Passing')
```

### Indirect effect (kind-of) the old fashioned way

#### Effect of working on number of hours studied

```{r ieModel}
# indirect effect model
indirect_effect1 <- stan_glm(study ~ work, 
                         family = 'gaussian',
                         seed = 938,
                         data = grades)
```

```{r iePlotPosterior, fig.height = 3, fig.width = 4}
indirect_effect1 %>%
  add_fitted_draws(newdata = data.frame(work = work_order), seed = 1837) %>%
  two_group_plot('Effect of Working on Hours Studied') +
  scale_x_continuous() +
  xlab('Hours worked')
```

#### Effect of number of hours studied on working

```{r ieModel2}
indirect_effect2 <- stan_glm(grade ~ study + work, 
                         family = binomial(link = 'logit'),
                         seed = 938,
                         data = grades)
```

```{r iePlot2, fig.height = 3.5, fig.width = 3.5}
ggpredict(indirect_effect2, c("study [10:35]", "work"), seed = 39) %>% 
  plot() +
  scale_color_brewer(palette = "Set2", type = "Qual", labels = c("Not working", "Working")) +
  labs(title = 'Probability of Passing Class',
       x = 'Hours studied',
       y = 'Probability of passing',
       color = NULL) +
  theme(legend.position = 'bottom',
        plot.title = element_text(size = 10))
```

#### Putting it all together: Indirect effect of working on passing class

To calculate, multiply the two coefficients.
Issue, the hours studied coefficient will be in log-odds since it is a logistic regression model. 

```{r}
# posterior draws of work to study model
ie1_post_pred <- indirect_effect1 %>%
  add_fitted_draws(newdata = data.frame(work = work_order), n = 300, seed = 893) %>%
  ungroup() %>%
  select(work_original = work, study = .value) %>%
  mutate(work = 1)

kable(head(ie1_post_pred, 10), digits = 2)
```

```{r}
indirect_effect_post <- indirect_effect2 %>%
  add_fitted_draws(newdata = ie1_post_pred, n = 300, seed = 76) %>%
  ungroup()

ie2_post_pred0 <- indirect_effect_post %>%
  filter(work_original == 0) %>%
  mutate(.row = .row - 300) %>%
  select(work_study = study, .row, .draw, value_work = .value)

ie2_post_pred1 <- indirect_effect_post %>%
  filter(work_original == 1) %>%
  select(notwork_study = study, .row, .draw, value_notwork = .value)

indirect_effect_post_diff <- ie2_post_pred0 %>%
  left_join(ie2_post_pred1, by = c('.row', '.draw')) %>%
  mutate(diff = value_notwork - value_work,
         type_effect = 'Indirect Effect')
```


```{r fig.height = 3, fig.width = 4}
indirect_effect_post %>%
  select(work = work_original, .value) %>%
  two_group_plot('Probability of Passing Class by Work Status')
```

```{r fig.height = 2.5, fig.width = 4}
posterior_diff_plot(indirect_effect_post_diff, x_limits = c(0, .25), plot_title = 'Indirect Effect of Working on Passing')
```

### Compare results to `mediation` package

```{r}
med_fit <- lm(study ~ work, data = grades)

out_fit <- glm(grade ~ work + study,
              data = grades, family = binomial("logit"))

med_out <- mediate(med_fit, out_fit, treat = "work", mediator = "study",
                  robustSE = TRUE, sims = 500)

# get estimates and cnof. intervals for frequentist effect size estimates with mediation package

# dataframe of confidence intervals, with two columns (one for lower and upper bounds)
# will be used for plotting effect sizes
upper_lower_ci <- bind_rows(med_out$z.avg.ci, med_out$d.avg.ci, med_out$tau.ci)

mediation_effects <- tibble(
  effect_type = c('Direct Effect', 'Indirect Effect', 'Total Effect'),
  effect_size = c(med_out$z.avg, med_out$d.avg, med_out$tau.coef),
  lower_ci = upper_lower_ci[[1]],
  upper_ci = upper_lower_ci[[2]],
  effect_group = 'Frequentist (mediation package)'
)

# get 95% credible intervals for bayesian effects
bayes_effect <- bind_rows(list(direct_effect_post, indirect_effect_post_diff, total_effect_post)) %>%
  group_by(type_effect) %>%
  median_hdi(diff)
```

```{r raceMediation, fig.cap = "Direct, indirect, and total effects of working on class passage. Values are the average percentage point difference in class passage.", fig.height = 3.5, fig.width = 5}
bayes_effect %>%
  select(effect_type = type_effect, effect_size = diff, lower_ci = .lower, upper_ci = .upper) %>%
  mutate(effect_group = "Bayesian Mediation") %>%
  bind_rows(mediation_effects) %>%
    ggplot(aes(effect_size, fct_rev(effect_type), color = effect_group)) +
    geom_point() +
    geom_segment(aes(x = lower_ci, xend = upper_ci, yend = effect_type)) +
    geom_vline(xintercept = 0, alpha = .6) +
    scale_color_brewer(palette = "Set2", type = "Qual") +
    scale_x_continuous(labels = scales::percent_format(accuracy = 1)) +
    facet_wrap(~effect_group, ncol = 1) +
    labs(title = 'Cumulative Effects of Race on Bar Passage',
           x = "Percentage point difference (Non-workers minus Workers)\nIn avg. probability of class passage",
           y = NULL) +
    theme(legend.position = 'none')
```

## The versitility of mediation analysis